# Plotting 

In this chapter, we will go over some _Time Series_ examples.
The aim of this chapter is to help you grasp some of the ideas about plotting.

## Example Data

Click [OSU_enroll.csv](http://www.chrisbilder.com/stat878/sections/2/OSU_enroll.csv) to download data.

```{r}
osu.enroll <- read.csv(file = "OSU_enroll.csv", 
    stringsAsFactors = TRUE)
```

```{r}
head(osu.enroll)
```

```{r}
tail(osu.enroll)
```

```{r}
x <- osu.enroll$Enrollment
```


```{r}
#One way to do plot
dev.new(width = 8, height = 6, pointsize = 10) 

# we did not specify y-axis and R put our x in y-axis, time in  x-axis

plot(x = x, ylab = "OSU Enrollment", 
       xlab = "t (time)", type="l", col = "red", 
       main = "OSU Enrollment from Fall 1989 to Fall 2002", 
       panel.first = grid(col = "gray", lty = "dotted"))
points(x = x, pch = 20, col = "blue")



```

```{r}
# A little different version of the plot
  plot(x = x, ylab = "OSU Enrollment", type = "o", xlab = "t (time)", col = "red",
    main = "OSU enrollment data", panel.first = grid(col = "gray", lty = "dotted"))
```


```{r}
dev.new(width = 8, height = 6, pointsize = 10) 

# we did not specify y-axis and R put our x in y-axis, time in  x-axis

plot(x = x, ylab = "OSU Enrollment", 
       xlab = "t (time)", type="l", col = "red", 
       main = "OSU Enrollment from Fall 1989 to Fall 2002", 
       panel.first = grid(col = "gray", lty = "dotted"))

points(x = osu.enroll$Enrollment, pch = 20, col = "blue")
```

Altenatively, you can do the same thing using ggplot.

```{r}
library(ggplot2)

# Create a data frame
df <- data.frame(osu.enroll)

# Create the plot
ggplot(df, aes(x = t, y = Enrollment)) +
  geom_line(colour = "red") +  # Line plot
  geom_point(shape = 20, colour = "blue") +  # Add points
  labs(x = "t (time)", y = "OSU Enrollment", 
       title = "OSU Enrollment from Fall 1989 to Fall 2002") +  # Set axis labels and title
  theme_bw() +  # Set the theme to a white background with black lines
  theme(panel.grid.major = element_line(colour = "gray", linetype = "dotted"))  # Add gray dotted lines to the plot
```

When only x is specified in the `plot()` function, R puts this on the y-axis and uses the observation number on the x-axis.

Compare this to the next plot below where both x and y arguments are specified.  

```{r}
#More complicated plot
fall <- osu.enroll[osu.enroll$Semester == "Fall",]
spring <- osu.enroll[osu.enroll$Semester == "Spring",]
summer <- osu.enroll[osu.enroll$Semester == "Summer",]

plot(y = fall$Enrollment, x = fall$t,
    ylab = "OSU Enrollment", xlab = "t (time)", 
    col = "blue", 
    main = "OSU Enrollment from Fall 1989 to Fall 2002", 
    panel.first = grid(col = "gray", lty = "dotted"), 
    pch = 1, type = "o", ylim = c(0,max(osu.enroll$Enrollment)))

lines(y = spring$Enrollment, x = spring$t, col = "red", 
    type = "o", pch = 2)

lines(y = summer$Enrollment, x = summer$t, col = 
    "darkgreen", type = "o", pch = 3)
    
legend(x="center", legend= c("Fall","Spring","Summer"), pch=c(1,2,3), lty=c(1,1,1), col=c("blue","red","darkgreen"), bty="n")


```

```{r}
#Another way to do plot with actual dates
plot(y = osu.enroll$Enrollment, 
    x = as.Date(osu.enroll$date, format = "%m/%d/%Y"), 
    xlab = "Time", type = "l", col = "red",  
    main = "OSU Enrollment from Fall 1989 to Fall 2002",
    ylab = "OSU Enrollment")

points(y = osu.enroll$Enrollment, 
    x = as.Date(osu.enroll$date, format = "%m/%d/%Y"), pch 
    = 20, col = "blue")

#Create own gridlines
# v specifies vertical line; h specifies horizontal line
 abline(v = as.Date(c("1990/1/1", "1992/1/1", "1994/1/1", 
    "1996/1/1", "1998/1/1", "2000/1/1", "2002/1/1")),
    lty = "dotted", col = "lightgray")
 abline(h = c(10000, 15000, 20000), lty = "dotted", col = 
    "lightgray")
     
```

```{r}
# Autocorrelation

rho.x <- acf(x = x, type = "correlation", main = "OSU Enrollment series")
rho.x
rho.x$acf[1:9]
  
```


## S&P500 Index

Click [SP500weekly.csv](http://www.chrisbilder.com/stat878/sections/2/SP500weekly.csv) to download data.

```{r}
SP500 <- read.csv(file="SP500weekly.csv",stringsAsFactors = TRUE)
```

```{r}
head(SP500)
```

```{r}
tail(SP500)
```

```{r}
x <- SP500$Close
```

```{r}
#One way to do plot
dev.new(width = 8, height = 6, pointsize = 10) 
#again, we do not specify y-axis here
plot(x = x, ylab = "S&P 500 Index", xlab = "t (time)", 
    type = "l", col = "red", main = "S&P 500 Index from 
    1/1/1995 to 10/25/2021 (weekly)", 
    panel.first = grid(col = "gray", lty = "dotted"))

```

```{r}
#Another way to do plot with actual dates
plot(y = x, x = as.Date(SP500$WeekStart, format =
    "%m/%d/%Y"), xlab = "Time", type = "l", col = "red", main 
    = "S&P 500 Index from 1/1/1995 to 10/25/2021 (weekly)",
    ylab = "S&P 500 Index")

#Create own gridlines
abline(v = as.Date(c("1995/1/1", "2000/1/1", "2005/1/1", 
    "2010/1/1", "2015/1/1", "2020/1/1")), lty = "dotted", 
    col = "lightgray")

abline(h = seq(from = 0, to = 5000, by = 1000), lty = 
    "dotted", col = "lightgray")

```

```{r}
# One more way with fine control of the dates
plot(y = x, x = as.Date(SP500$WeekStart, format = 
    "%m/%d/%Y"), xlab = "Time", type = "l", col = "red", 
    main = "S&P 500 Index from 1/1/1995 to 10/25/2021 
    (weekly)", ylab = "S&P 500 Index", xaxt = "n")

axis.Date(side = 1, at = seq(from = as.Date("1995/1/1"),
    to = as.Date("2021/12/31"), by = "years"), labels = 
    format(x = seq(from = as.Date("1995/1/1"), to = 
    as.Date("2021/12/31"), by = "years"), format = "%b%y"), 
    las = 2)  #las changes orientation of labels

#Create own gridlines
abline(v = as.Date(c("1995/1/1", "2000/1/1", "2005/1/1", 
    "2010/1/1", "2015/1/1", "2020/1/1")), lty = "dotted", 
    col = "lightgray")
abline(h = seq(from = 0, to = 5000, by = 1000), lty = 
    "dotted", col = "lightgray")

```

## Sunspots

Click [SN_y_tot_V2.0.csv](http://www.chrisbilder.com/stat878/sections/2/SN_y_tot_V2.0.csv) to download data.

```{r}
sunspots <- read.table(file = "SN_y_tot_V2.0.csv", sep = 
    ";", col.names = c("Mid.year", "Mean.total", 
   "Mean.SD.total", "Numb.obs.used", "Definitive"))
```

```{r}
head(sunspots)
```

```{r}
tail(sunspots)
```

```{r}
dev.new(width = 8, height = 6, pointsize = 10)

#again, we did not specify y-axis here
plot(x = sunspots$Mean.total, ylab = "Number of 
    sunspots", xlab = "t (time)", type = "l", col = "red", 
    main = "Sunspots per year from 1700 to 2020",
    panel.first = grid(col = "gray", lty = "dotted"))

points(x = sunspots$Mean.total, pch = 20, col = "blue")


```

```{r}
# Include dates
plot(y = sunspots$Mean.total, x = sunspots$Mid.year, ylab 
    = "Number of sunspots", xlab = "Year", type = "l", col 
    = "red", main = "Sunspots per year from 1700 to 2020",
    panel.first = grid(col = "gray", lty = "dotted"))

points(y = sunspots$Mean.total, x = sunspots$Mid.year, 
    pch = 20, col = "blue")

```

```{r}
#Convert to an object of class "ts"

x <- ts(data = sunspots$Mean.total, start = 1700, frequency 
    = 1)

x

```

```{r}
class(x)

class(sunspots$Mean.total)
```

### plot.ts()

plot() is a generic function - uses the plot.ts() method function

```{r}
# we did not specify y-axis here, but x is now ts
plot(x = x, ylab = expression(paste(x[t], " (Number of 
   sunspots)")), xlab = "Year", type = "o", col = "red", main 
   = "Sunspots per year from 1700 to 2020")
```

```{r}
plot.ts(x = x, ylab = expression(paste(x[t], " (Number of sunspots)")),
  xlab = "Year", type = "o", col = "red", main = "Sunspots per year from 1700 to 2020")


#type = "b" also works for "both" points and lines, but it leaves spaces between the points and lines
```

