# Integrated Models for Nonstationary Data

We have assumed that $x_t$ is stationary so far. What if $x_t$ is not stationary in the mean? Use differencing to make a transformation stationary in the mean.  

**Autoregressive Integrated Moving Average model**

- ARIMA(p,d,q) – p is AR order, d is differencing order, and q is MA order
- $\phi(B)(1-B)^dx_t = \theta(B)w_t$ where $w_t \sim ind. N(0,\sigma_w^2)$
- Let $(1-B)^dx_t = y_t.$ Then $\phi(B)y_t = \theta(B)w_t$ is an ARMA(p,q) model.  
- The “integrated” name results from transforming back from the $y_t$ to $x_t$ by “integrating” (put together) or “summing” the $y_t$’s. With first differences, we have $$\sum_{k=1}^{t}y_k=\sum_{k=1}^{t}x_k-x_{k-1}=x_t-x_0=x_t$$ if $x_0=0$

 
Because higher order differencing is the result of continuing to apply first differencing, the same process would be done in those cases.    

:::{.example}

**ARIMA(1,1,1) with $\phi_1=0.7, \theta_1=0.4, \sigma_w^2=9, n=200$ (arima111_sim.R)**

$\phi(B)(1-B^d)x_t=\theta(B)w_t,$ where $w_t \sim ind.N(0.9)$

This can be rewritten as 

$$(1-\phi_1B)(1-B)x_t=(1+\theta_1B)w_t\\
\iff (1-\phi_1B)(x_t-x_{t-1})=(1+\theta_1B)w_t\\
\iff x_t-x_{t-1}-\phi_1x_{t-1}+\phi_1x_{t-2}=w_t+\theta_1w_{t-1}\\
\iff x_t=(1+\phi_1)x_{t-1}-\phi_1x_{t-2}+w_t+\theta_1w_{t-1}$$

Using the above representation with only $x_t$ on the left side, one could use the `for loop` to simulate observations from this model. Instead, one can use `arima.sim()` to do it as follows, 

```{r}
#Data could be simulated using the following code - notice the use of the order option.  

set.seed(6632)

x <- arima.sim(model = list(order=c(1,1,1), ar=0.7, ma=0.4), n=200, rand.gen=rnorm, sd=3)
```

Notice the addition of the `order` option to specify p, d, and q.  

I had already simulated observations from the model in the past and put them in the comma delimited file. I am going to use this data for the rest of the example.  

```{r}
arima111 <- read.csv(file = "arima111.csv")
head(arima111)
tail(arima111)
```

```{r}
x <- arima111$x
```

```{r}
#Plot of the data
# dev.new(width = 8, height = 6, pointsize = 10)  
par(mfrow = c(1,1))

plot(x = x, ylab = expression(x[t]), xlab = "t", type = 
   "l", col = "red", main = expression(paste("ARIMA 
   model: ", (1 - 0.7*B)*(1-B)*x[t] == (1 + 0.4*B)*w[t])), 
   panel.first = grid(col = "gray", lty = "dotted"))

points(x = x, pch = 20, col = "blue")

```

```{r}
#ACF and PACF of x_t
#dev.new(width = 8, height = 6, pointsize = 10)
par(mfcol = c(2,3))

acf(x = x, type = "correlation", lag.max = 20, ylim = 
     c(-1,1), main = expression(paste("Estimated ACF plot 
     for ", x[t])))

pacf(x = x, lag.max = 20, ylim = c(-1,1), xlab = "h", 
     main = expression(paste("Estimated PACF plot for ", 
      x[t])))
 
#ACF and PACF of first differences
acf(x = diff(x = x, lag = 1, differences = 1), type = 
      "correlation", lag.max = 20, ylim = c(-1,1), main = 
      expression(paste("Estimated ACF plot for ", x[t],"-",x[t-1])))

pacf(x = diff(x = x, lag = 1, differences = 1), lag.max 
      = 20, ylim = c(-1,1), xlab = "h", main = 
      expression(paste("Estimated PACF plot for ", x[t],"-",x[t-1])))
 
#True ACF and PACF for ARIMA(1,0,1) (without differences)

plot(y = ARMAacf(ar = 0.7, ma = 0.4, lag.max = 20), x = 
      0:20, type = "h", ylim = c(-1,1), xlab = "h", ylab = 
      expression(rho(h)), main = "True ACF for 
      ARIMA(1,0,1)")

abline(h = 0)

plot(x = ARMAacf(ar = 0.7, ma = 0.4, lag.max = 20, pacf 
      = TRUE), type = "h", ylim = c(-1,1), xlab = "h", ylab 
      = expression(phi1[hh]), main = "True ACF for 
      ARIMA(1,0,1)")
abline(h = 0)

```

```{r}
#Plot of the first differences
#dev.new(width = 8, height = 6, pointsize = 10)  
par(mfrow = c(1,1))

plot(x = diff(x = x, lag = 1, differences = 1), ylab = 
     expression(x[t] - x[t-1]), xlab = "t", type = "l", col 
      = "red", main =  "Plot of data after first 
      differences", panel.first = grid(col = "gray", lty = 
     "dotted"))
points(x = diff(x = x, lag = 1, differences = 1), pch = 
     20, col = "blue")

```

:::


Notes:

- The $x_t vs. t$ plot shows characteristics of a nonstationary in the mean time series.  
- The ACF plot shows very large autocorrelations.  Remember that this is a characteristic of a nonstationary in the mean time series.  
- After first differences, the ACF and PACF look like the ACF and PACF from an ARMA(1, 1) with $\phi_1 = 0.7$ and $\theta_1 = 0.4$. The plot of the first differences themselves now look like a sample from a stationary process.  
- While we have not talked about how to estimate model parameters, we can still take a quick look at what if the parameters are estimated. The `arima()` function in R can do it.  

```{r}
# Fit model

mod.fit <- arima(x=x, order=c(1,1,1))
mod.fit
```
These estimates are relatively close to the values used in `arima.sim()`!

```{r}
#Covariance matrix
  mod.fit$var.coef
```

```{r}
 #Test statistic for Ho: phi1 = 0 vs. Ha: phi1 ≠ 0
  z <- mod.fit$coef[1]/sqrt(mod.fit$var.coef[1,1])
  z
  2*(1-pnorm(q = z, mean = 0, sd = 1))

  #Test statistic for Ho: theta1 = 0 vs. Ha: theta1 <> 0
  z <- mod.fit$coef[2]/sqrt(mod.fit$var.coef[2,2])
  z
  2*(1-pnorm(q = abs(z), mean = 0, sd = 1))

  # confidence interval
  confint(object = mod.fit, level = 0.95)
```

```{r}
#Shows how to use the xreg option
# dev.new(width = 8, height = 6, pointsize = 10)
arima(x = x, order = c(1, 1, 1), xreg = rep(x = 1, times = length(x)))
```
