# Drift Term

## Random Walk Model

A random walk model is simply an ARIMA(0,1,0):

$(1-B)x_t=w_t\implies x_t=x_{t-1}+w_t$

where $w_t \sim ind.(0, \sigma_w^2)$

The name comes about through $x_t$ being $x_{t-1}$ plus a random “movement”.  

A drift term $\delta$ can be added to the model,

$x_t=\delta+x_{t-1}+w_t$

This can be rewritten then as a cumulative sum, 

$$x_t=\delta+(\delta+x_{t-2}+w_{t-1})+w_t\\
=2\delta+x_{t-2}+w_{t-1}+w_t\\
=2\delta+(\delta+x_{t-3}+w_{t-2})+w_{t-1}+w_t\\
=3\delta+x_{t-3}+w_{t-2}+w_{t-1}+w_t\\
\vdots\\
x_t=\delta t+\sum_{j=1}^{t}w_j$$

One can see that the drift term creates a non-stochastic trend in the series (i.e., allows $x_t$ to drift away from 0). 

Notice the random walk model with a drift term is not stationary because the mean is a function of time.  

$E(x_t)=\delta t +\sum_{j=1}^{t}E(w_j)=\delta t$

## Estimating the drift term

When `arima()` is used with differencing, the `include.mean` argument is set to `FALSE`. This is because differencing removes the trend and allows for the differenced series to have a mean of 0. Note that even if you say `include.mean = TRUE`, there will be no “mean” estimated. Here’s what R’s `arima()` help says for the include.mean argument,

> `include.mean`	Should the ARIMA model include a mean term? The default is TRUE for undifferenced series, FALSE for differenced ones (where a mean would not affect the fit nor predictions).

As shown in the random walk model example, there may be times where you do want to estimate a drift term, like $\delta$, still. This leads to an ARIMA model with the drift term, 

$$(1-\phi_1B-\phi_2B^2-...-\phi_pB^p)(1-B)^dx_t=\\
\delta+(1+\theta_1B+\theta_2B^2+...+\theta_qB^q)w_t\\
\iff \phi(B)(1-B)^dx_t=\delta+\theta(B)w_t$$

Most time series textbooks will not include the constant term in the model when there is differencing. For example, Wei’s textbook says (referring to $\delta$ as $\theta_0$),

> We assume that $\theta_0 = 0$, unless it is clear from the data or the nature of the problem that a deterministic [non-stochastic] trend is really needed.  

This is why R does not include the possible estimation of $\delta$ in the `arima()` function.  

If you did want to include $\delta$, there are a few ways to estimate it. 

1. Use `xreg = 1:length(x) `in `arima()` where x contains the data used with the `arima()`.   
2. Use the `sarima()` and `sarima.for()` functions in the `astsa` package for fitting the model and forecasting, respectively. The `sarima()` name comes from an extension of the ARIMA model that will be discussed later in our course to allow for seasonality. 
3. Do all differencing needed BEFORE invoking `arima()`. Use `d = 0` in `arima()` and `include.mean = TRUE`. 

Below are examples of their implementations. Shumway and Stoffer also include an example when finding the best model for their GNP data.  

:::{.example}

**ARIMA(1,1,1) with $\phi_1=0.7, \theta_1=0.4, \sigma_w^2=9, n=200$ (arima111_sim.R)**

Of course, $\delta$ should be 0 for this data set since the data were simulated directly from an ARIMA(1,1,1). However, we can still investigate what would happen if $\delta$ was estimated. 

First, the original model’s fit has been reproduced below.  

```{r}
arima111 <- read.csv(file = "arima111.csv")
x <- arima111$x
```

```{r}
mod.fit <- arima(x = x, order = c(1, 1, 1))
mod.fit
```

```{r}
#Forecasts 5 time periods into the future
fore.mod <- predict(object = mod.fit, n.ahead = 5, se.fit 
                      = TRUE) 
fore.mod
```

The estimated model is 

$(1 - 0.6720B)(1 - B)x_t = (1 + 0.4681B)w_t$ with $\hat \sigma_w^2= 9.56$

Next, the models are fit including an estimate of $\delta$.  

1. Using `xreg = 1:length(x)` in `arima()` where x contains the data

```{r}
mod.fit2 <- arima(x = x, order = c(1, 1, 1), xreg = 
    1:length(x))

mod.fit2

```

$\hat \delta=-1.6847, \sqrt{\hat Var(\hat \delta)}=0.8815, \hat \sigma_w^2=9.404$

```{r}
fore.mod2 <- predict(object = mod.fit2, n.ahead = 5, 
    se.fit = TRUE, newxreg = (length(x)+1):(length(x)+5))
fore.mod2
```

Notice in our example,` newxreg = (length(x)+1):(length(x)+5))` basically means t=201,...,205.

$H_0: \delta=0\quad vs. \quad  H_a:  \delta \ne 0$

Wald statistics: $\frac{\hat \delta-0}{\sqrt{\hat Var(\hat \delta)}}=\frac{-1.6847}{0.8815}$

```{r}
# calculating wald statistics for hypothesis testing
mod.fit2$coef[3]/sqrt(mod.fit2$var.coef[3,3])

# surprisingly, we see delta ≠ 0 statistically significant
```

```{r}
1:length(x) 
```


Note the `newxreg` option used in `predict()` and the values that I put into it. The estimated model is 

$(1 - 0.6382B)(1 - B)x_t = -1.6847 + (1 + 0.4826B)w_t$ with $\hat \sigma_w^2=9.404$

A Wald test was performed for Ho:$\delta$ = 0 vs. Ha:$\delta \ne$ 0, and, surprisingly, $\delta$ is marginally significant. But we simulate our data with $\delta=0$. This is an example of a type I error!  

2. The `astsa` package 

```{r}
library(astsa)
mod.fit.SS <- sarima(xdata = x, p = 1, d = 1, q = 1)
mod.fit.SS

```

```{r}
names(mod.fit.SS)
```

```{r}
mod.fit.SS$fit$coef
```

```{r}
mod.fit.SS$fit$var.coef
```

The plot that includes information about the residuals will be discussed later in the course. The estimated model is 

$(1 - 0.6382B)(1 - B)x_t = -1.6847 + (1 + 0.4826B)w_t$ with $\hat \sigma_w^2=9.404$

The forecasts are shown below. Use `sarima.for` to forecast.

```{r}
save.for <- sarima.for(xdata = x, n.ahead = 5, p = 1, d 
    = 1, q = 1)  
save.for
```

The envelope surrounding the forecasts represent $\pm 1$ and $\pm 2$ multiplied by  $\sqrt{Var(x_{n+m}-\tilde x^n_{n+m})}$

3. Perform the differencing BEFORE invoking `arima()`, Use `d = 0` in `arima()`, and `include.mean = TRUE`. 

See the program for partial implementation. Methods #1 and #2 are easier, so these will be used in this course.   

If you use method 3, you will forecast $v_t=(1-B)^dx_t$, so your forecast is not in the scale of $x_t$. This can be cumbersome.

```{r}
#Do all differencing needed BEFORE invoking arima(). Use d = 0 in arima() and include.mean = TRUE.

x.diff <- diff(x = x, lag = 1, differences = 1)
mod.fit.diff <- arima(x = x.diff, order = c(1, 0, 1))
mod.fit.diff

```


```{r}
fore.mod.diff <- predict(object = mod.fit.diff, n.ahead = 5, se.fit = TRUE)
fore.mod.diff

  #Problem with doing the above method is the need to get the undifferenced data back
  #  Would need to do "integrating"
```


:::


